Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-09 ind_files/99_Appendix_D_examples.tex
Fix in-text citation spacing and paragraph spacing

The deterministic trajectory \eqref{ex_1D_stoch_logistic_det_limit} has two fixed points, one at $x=0$ (extinction) and one at $x=1$ (corresponding to a population size of $n=K$). For $r > 0$, $x=0$ is unstable and $x=1$ is a global attractor, meaning in the deterministic limit, when $r > 0$, all populations end up at $x=1$ given enough time. The stochastic dynamics \eqref{ex_1D_stoch_logistic_full_SDE} and \eqref{ex_1D_stoch_logistic_WNA} depend not only on $r$, but also on $v$, the sum of the birth and death rates. It has been proven that $X_t = 0$ is the only recurrent state for the full stochastic dynamics \eqref{ex_1D_stoch_logistic_full_SDE}, meaning that every population is guaranteed to go extinct\footnote{This can be proven using tools from Markov chain theory. For those interested, the proof uses ergodicity to arrive at a contradiction if any state other than $0$ exhibits a non-zero density at steady state.} given enough time \citep{nasell_extinction_2001}, thus illustrating an important difference between finite and infinite populations. $X_t = 0$ is also an `absorbing' state since once a population goes extinct, it has no way of being revived in this model. However, if $K$ is large enough, the eventual extinction of the population may take a very long time. In fact, we can make the expected time to extinction arbitrarily long by making $K$ sufficiently large. Thus, for moderately large values of $K$, it is biologically meaningful only to look at a weaker version of the steady state distribution by imposing the condition that the population does not go extinct and looking at the `transient' dynamics \citep{hastings_transients_2004}. Conditioned on non-extinction, the solution to \eqref{ex_1D_stoch_logistic_full_SDE} has a `quasistationary' distribution about the deterministic attractor $X_t = 1$, with some variance reflecting the effect of noise-induced fluctuations in population size \citep{nasell_extinction_2001} due to the finite size of the population. The weak-noise approximation \eqref{ex_1D_stoch_logistic_WNA} implicitly assumes non-extinction by only measuring small fluctuations from the deterministic solution to \eqref{ex_1D_stoch_logistic_det_limit} and thus, at steady state, naturally describes a quasistationary distribution centered about $X_t = 1$. The steady-state density (probability density function as $t \to \infty$) of the exact birth-death process \eqref{ex_1D_stoch_logistic_BD_eqns} is compared with that predicted by \eqref{ex_1D_stoch_logistic_full_SDE} and \eqref{ex_1D_stoch_logistic_WNA} for various values of $K$ in figure \ref{fig_1D_stoch_logistic_densities}.

@@

The methods outlined in the above section have recently been used to study the population dynamics of a finite population playing a so-called `matrix game' (An evolutionary game for which you can write down a payoff matrix) with 2 pure strategies \citep{tao_stochastic_2007}. Based on the interpretation of what each type represents, this is mathematically equivalent to studying frequency-dependent selection on a one-locus two-allele gene (with a bijective genotype-phenotype map and no mutations) or studying two-species competitive Lotka-Volterra dynamics, as we will show below. The stochastic Lotka-Volterra competition model shown below has also been proved to be equivalent to an $m$-allele Moran model under certain limits \citep{constable_mapping_2017}.\\
\\
Let us imagine a population with $m$ types of individuals that are interacting according to some ecological rules. Let the state of the population be characterized by the vector $\mathbf{v}(t) = [v_1(t),v_2(t),\ldots,v_m(t)]$, where $v_i(t)$ is the number of type $i$ individuals at time $t$. Let the birth and death rates of the $i$th type be given by:
\label{nD_example_numbers_b_d_rates}
b_i(\mathbf{v}) &= \left(\lambda + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\beta_{ij}v_j\right)\right)v_i\\
d_i(\mathbf{v}) &= \left(\mu + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\delta_{ij}v_j\right)\right)v_i
where $K > 0$ is our system size parameter (and represents a global carrying capacity across all types), $\lambda > 0$ and $\mu > 0$ are suitable positive constants representing the baseline natality and mortality common to all types, and $\beta_{ij}$ and $\delta_{ij}$ are constants describing the effect of type $j$ individuals on the birth and death rate of type $i$ individuals respectively. The sign of $M_{ij} \coloneqq \beta_{ij} - \delta_{ij}$ determines whether type $j$ has a net positive or negative effect on the growth of type $i$. In ecological communities, this is a per-capita ecological interaction effect. In game-theoretic terms, we can interpret $M_{ij}$ as the payoff obtained by a type $j$ individual playing against a type $i$ individual. I assume that $| M_{ij} | \ll K$. The values $M_{ij}$ are often collected in an $m \times m$ matrix $\mathbf{M}$ called the `payoff matrix' (in evolutionary game theory) or `interaction matrix' (in Lotka-Volterra models). Lotka-Volterra models also frequently assume that the diagonal elements $M_{ii}$ are all equal, though I will not make that assumption here.\\
\\
Going from population numbers $\mathbf{v}$ to densities $\mathbf{x} = \mathbf{v}/K$, we obtain the birth and death rates:
\label{nD_example_density_b_d_rates}
b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i
Thus, we have
A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)
Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
and
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(\nu + \sum\limits_{j=1}^{m}T_{ij}{(\mathbf{X}_{t})}_j) 
From \eqref{nD_det_limit}, we see that the deterministic limit is a set of $m$ coupled ODEs given by
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)
These are precisely the Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\label{nD_example_fitness}
w_i(\mathbf{x}) = r + \sum\limits_{j=1}^{m}M_{ij}x_j
If $p_i(t)$ is the frequency of type $i$ individuals in the population at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics \citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to \citep{constable_stochastic_2013} and \citep{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.\\
Let the solution to the equations \eqref{nD_example_det_limit} be given by $\mathbf{a}(t) = [a_1(t),\ldots,a_m(t)]$. For the weak noise approximation, we can Taylor expand $A^{\pm}_i$ and use \eqref{nD_WNA_directional_derivative_for_replicator_eqns} to compute the directional derivative as:
D_i &= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial w_i}{\partial x_k}\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial}{\partial x_k}(r+\sum\limits_{j=1}^{m}M_{ij}x_j)\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_kM_{ik}\\
\Rightarrow D_i &= y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\label{nD_example_directional_derivative}
where we have used the fact that $w_i(\mathbf{y}) = r + \sum\limits_{k=1}^{m}y_kM_{ik}$ (from \eqref{nD_example_fitness}) in the last step. Thus, in the weak noise approximation of our process, the dynamics are given by
\mathbf{x}(t) = \mathbf{a}(t) + \frac{1}{\sqrt{K}}\mathbf{y}(t)
where the stochastic fluctuations $\mathbf{y}(t)$ satisfy the linear Fokker-Planck equation
\resizebox{1.1\textwidth}{!}{$\displaystyle\frac{\partial P}{\partial t}(\mathbf{y},t) = \sum\limits_{i=1}^{m}\left(-\frac{\partial}{\partial y_i}\left\{\left(y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\right)P(\mathbf{y},t)\right\}+\frac{1}{2}\left(a_i\left(\nu + \sum\limits_{j=1}^{m}T_{ij}a_j\right)\right)\frac{\partial^2}{\partial{y_i}^2}P(\mathbf{y},t)\right)$}
Using \eqref{nD_example_directional_derivative} in \eqref{nD_moment_eqn_mean}, we see that the fluctuations are expected to evolve as:
\label{nD_example_moment_eqn_mean}
\frac{d}{dt}\mathbb{E}[y_i] = w_i(\mathbf{a})\mathbb{E}[y_i] + a_i\sum\limits_{k=1}^{m}M_{ik}\mathbb{E}[y_k]
or, in matrix form:
\resizebox{0.93\textwidth}{!}{$\displaystyle
	\frac{d}{dt}\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	=
	\begin{bmatrix}
	(r + \sum\limits_{j=1}^{m}M_{1j}a_j + a_1M_{11}) & a_1M_{12} & a_1M_{13} & \dots & \dots & \dots & a_1M_{1m}\\
	a_2M_{21} & (r + \sum\limits_{j=1}^{m}M_{2j}a_j + a_2M_{22}) & a_2M_{23} & \dots & \dots & \dots & a_2M_{2m}\\
	\vdots &  & \ddots & &  & & \vdots\\
	a_{i}M_{i1} & a_iM_{i2} & a_iM_{i3} & \dots & (r + \sum\limits_{j=1}^{m}M_{ij}a_j + a_iM_{ii}) & \dots & a_iM_{im}\\
	\vdots &  &  & & & \ddots & \vdots\\
	a_mM_{m1} & a_mM_{m2} & a_mM_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}M_{mj}a_j + a_mM_{mm})
	\end{bmatrix}
	\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	$}
The eigenvalues of the first matrix on the RHS will tell us whether the fixed point $\mathbb{E}[y_i] = 0 \ \forall \ i$ (the only fixed point of this system) is stable, or whether fluctuations are expected to grow (up to the point where the fluctuations are so large that the WNA is no longer valid). In the $m=2$ case, \cite{tao_stochastic_2007} have shown that $\mathbb{E}[y_i] = 0 \ \forall \ i$ is a stable fixed point for this system iff the point $\mathbf{y}$ is an ESS (in the usual game-theoretic sense) for the matrix game defined by the payoff matrix $\mathbf{M}$.

@@

In Chapter \ref{chap_infD_processes}, we used various approximations to arrive at the linear functional Fokker-Planck equation
\label{functional_WNE_chap_4}
    \frac{\partial P}{\partial t}(\zeta,t) = \int\limits_{\mathcal{T}}\left(-\frac{\delta}{\delta \zeta(x)}\left\{\mathcal{D}_{\zeta}[\mathcal{A}^{-}](x)P(\zeta,t)\right\}+\frac{1}{2}\mathcal{A}^{+}(x|\psi)\frac{\delta^2}{\delta\zeta(x)^2}\{P(\zeta,t)\}\right)dx
for describing stochastic fluctuations $\zeta$ from the deterministic solution obtained by solving \eqref{deterministic_traj}. Our goal is now to find a method to effectively detect and describe evolutionary branches (modes in trait space, corresponding to individual morphs) for this process. Following the methods used by Tim Rogers and colleagues for various special cases \citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015}, we will do this in a general manner by measuring the autocorrelation of the distribution of the population over trait space, a task made easier by moving to Fourier space.
\myfig{0.8}{Media/4.3_fourier_fig.png}{\textbf{Schematic description of Fourier analysis}. A function $\phi(x)$ (shown in red) over the trait space can be decomposed as the sum of infinitely many Fourier modes (shown in blue) $\phi_k$. In the Fourier dual space, we can look at the peaks of each of these Fourier modes: The magnitude of $\phi_k$ tells us how much it contributes to the actual function of interest $\phi$.}{fig_Fourier}
Specifically, a convenient theorem due to Weiner and Khinchin relates the autocorrelation of a probability distribution to its power spectral density via Fourier transformation. This has been extensively used in spatial ecology, and we too will make use of it here by carrying out a basis expansion of our functions in the Fourier basis $\{e^{ikx}\}_{k\in\mathbb{Z}}$. If $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ is a translation-invariant\footnote{This is horrible nomenclature by the mathematicians. Though `invariant' is the conventional name for this concept, the intended meaning is not really invariant but `equivariant'. Formally, let $\mathcal{F}$ be a suitable function space of real valued functions. For any $c \in \mathbb{R}$, let $T_c: \mathcal{F} \to \mathcal{F}$ be the translation operator on this space, defined by $T_c[f(x)] = f(x+c)$. An operator $L: \mathcal{F} \to \mathcal{F}$ is said to be translation-invariant if it commutes with $T_c$ for every $c \in \mathbb{R}$, \emph{i.e.} $T_c[L[f]] = L[T_c[f]] \ \forall \ f \in \mathcal{F} \ \forall \ c \in \mathbb{R}$} linear operator, then $\exp(ikx)$ acts as an eigenfunction, significantly simplifying the calculations. We therefore assume this below.\\
\\
Assume that $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ takes the form:
 \mathcal{D}_{\zeta}[\mathcal{A}^-](x,t) = L[\zeta(x,t)]   
for a translation-invariant linear operator $L$ that only depends on $x$ and $t$. The presence of phenotypic clustering and polymorphisms can be analyzed by examining the power spectrum of $\Tilde{P}_{0}(\zeta,s)$ over the trait space.\\

>>>>

The deterministic trajectory \eqref{ex_1D_stoch_logistic_det_limit} has two fixed points, one at $x=0$ (extinction) and one at $x=1$ (corresponding to a population size of $n=K$). For $r > 0$, $x=0$ is unstable and $x=1$ is a global attractor, meaning in the deterministic limit, when $r > 0$, all populations end up at $x=1$ given enough time. The stochastic dynamics \eqref{ex_1D_stoch_logistic_full_SDE} and \eqref{ex_1D_stoch_logistic_WNA} depend not only on $r$, but also on $v$, the sum of the birth and death rates. It has been proven that $X_t = 0$ is the only recurrent state for the full stochastic dynamics \eqref{ex_1D_stoch_logistic_full_SDE}, meaning that every population is guaranteed to go extinct\footnote{This can be proven using tools from Markov chain theory. For those interested, the proof uses ergodicity to arrive at a contradiction if any state other than $0$ exhibits a non-zero density at steady state.} given enough time~\citep{nasell_extinction_2001}, thus illustrating an important difference between finite and infinite populations. $X_t = 0$ is also an `absorbing' state since once a population goes extinct, it has no way of being revived in this model. However, if $K$ is large enough, the eventual extinction of the population may take a very long time. In fact, we can make the expected time to extinction arbitrarily long by making $K$ sufficiently large. Thus, for moderately large values of $K$, it is biologically meaningful only to look at a weaker version of the steady state distribution by imposing the condition that the population does not go extinct and looking at the `transient' dynamics~\citep{hastings_transients_2004}. Conditioned on non-extinction, the solution to \eqref{ex_1D_stoch_logistic_full_SDE} has a `quasistationary' distribution about the deterministic attractor $X_t = 1$, with some variance reflecting the effect of noise-induced fluctuations in population size~\citep{nasell_extinction_2001} due to the finite size of the population. The weak-noise approximation \eqref{ex_1D_stoch_logistic_WNA} implicitly assumes non-extinction by only measuring small fluctuations from the deterministic solution to \eqref{ex_1D_stoch_logistic_det_limit} and thus, at steady state, naturally describes a quasistationary distribution centered about $X_t = 1$. The steady-state density (probability density function as $t \to \infty$) of the exact birth-death process \eqref{ex_1D_stoch_logistic_BD_eqns} is compared with that predicted by \eqref{ex_1D_stoch_logistic_full_SDE} and \eqref{ex_1D_stoch_logistic_WNA} for various values of $K$ in figure \ref{fig_1D_stoch_logistic_densities}.

@@

The methods outlined in the above section have recently been used to study the population dynamics of a finite population playing a so-called `matrix game' (An evolutionary game for which you can write down a payoff matrix) with 2 pure strategies~\citep{tao_stochastic_2007}. Based on the interpretation of what each type represents, this is mathematically equivalent to studying frequency-dependent selection on a one-locus two-allele gene (with a bijective genotype-phenotype map and no mutations) or studying two-species competitive Lotka-Volterra dynamics, as we will show below. The stochastic Lotka-Volterra competition model shown below has also been proved to be equivalent to an $m$-allele Moran model under certain limits~\citep{constable_mapping_2017}.

Let us imagine a population with $m$ types of individuals that are interacting according to some ecological rules. Let the state of the population be characterized by the vector $\mathbf{v}(t) = [v_1(t),v_2(t),\ldots,v_m(t)]$, where $v_i(t)$ is the number of type $i$ individuals at time $t$. Let the birth and death rates of the $i$th type be given by:
\begin{equation}
\label{nD_example_numbers_b_d_rates}
\begin{aligned}
b_i(\mathbf{v}) &= \left(\lambda + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\beta_{ij}v_j\right)\right)v_i\\
d_i(\mathbf{v}) &= \left(\mu + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\delta_{ij}v_j\right)\right)v_i
\end{aligned}
\end{equation}
where $K > 0$ is our system size parameter (and represents a global carrying capacity across all types), $\lambda > 0$ and $\mu > 0$ are suitable positive constants representing the baseline natality and mortality common to all types, and $\beta_{ij}$ and $\delta_{ij}$ are constants describing the effect of type $j$ individuals on the birth and death rate of type $i$ individuals respectively. The sign of $M_{ij} \coloneqq \beta_{ij} - \delta_{ij}$ determines whether type $j$ has a net positive or negative effect on the growth of type $i$. In ecological communities, this is a per-capita ecological interaction effect. In game-theoretic terms, we can interpret $M_{ij}$ as the payoff obtained by a type $j$ individual playing against a type $i$ individual. I assume that $| M_{ij} | \ll K$. The values $M_{ij}$ are often collected in an $m \times m$ matrix $\mathbf{M}$ called the `payoff matrix' (in evolutionary game theory) or `interaction matrix' (in Lotka-Volterra models). Lotka-Volterra models also frequently assume that the diagonal elements $M_{ii}$ are all equal, though I will not make that assumption here.

Going from population numbers $\mathbf{v}$ to densities $\mathbf{x} = \mathbf{v}/K$, we obtain the birth and death rates:
\begin{equation}
\label{nD_example_density_b_d_rates}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i
\end{aligned}
\end{equation}
Thus, we have
\begin{equation*}
A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)
\end{equation*}
Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by
\begin{equation}
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\end{equation}
where 
\begin{equation*}
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
and
\begin{equation*}
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(\nu + \sum\limits_{j=1}^{m}T_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
From \eqref{nD_det_limit}, we see that the deterministic limit is a set of $m$ coupled ODEs given by
\begin{equation}
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{equation}
These are precisely the Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\begin{equation}
\label{nD_example_fitness}
w_i(\mathbf{x}) = r + \sum\limits_{j=1}^{m}M_{ij}x_j
\end{equation}
If $p_i(t)$ is the frequency of type $i$ individuals in the population at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\begin{align}
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{align}
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\begin{equation}
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   
\end{equation}
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics~\citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to~\citep{constable_stochastic_2013} and~\citep{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.

Let the solution to the equations \eqref{nD_example_det_limit} be given by $\mathbf{a}(t) = [a_1(t),\ldots,a_m(t)]$. For the weak noise approximation, we can Taylor expand $A^{\pm}_i$ and use \eqref{nD_WNA_directional_derivative_for_replicator_eqns} to compute the directional derivative as:
\begin{align}
D_i &= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial w_i}{\partial x_k}\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial}{\partial x_k}(r+\sum\limits_{j=1}^{m}M_{ij}x_j)\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_kM_{ik}\\
\Rightarrow D_i &= y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\label{nD_example_directional_derivative}
\end{align}
where we have used the fact that $w_i(\mathbf{y}) = r + \sum\limits_{k=1}^{m}y_kM_{ik}$ (from \eqref{nD_example_fitness}) in the last step. Thus, in the weak noise approximation of our process, the dynamics are given by
\begin{equation}
\mathbf{x}(t) = \mathbf{a}(t) + \frac{1}{\sqrt{K}}\mathbf{y}(t)
\end{equation}
where the stochastic fluctuations $\mathbf{y}(t)$ satisfy the linear Fokker-Planck equation
\begin{equation}
\resizebox{1.1\textwidth}{!}{$\displaystyle\frac{\partial P}{\partial t}(\mathbf{y},t) = \sum\limits_{i=1}^{m}\left(-\frac{\partial}{\partial y_i}\left\{\left(y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\right)P(\mathbf{y},t)\right\}+\frac{1}{2}\left(a_i\left(\nu + \sum\limits_{j=1}^{m}T_{ij}a_j\right)\right)\frac{\partial^2}{\partial{y_i}^2}P(\mathbf{y},t)\right)$}
\end{equation}
Using \eqref{nD_example_directional_derivative} in \eqref{nD_moment_eqn_mean}, we see that the fluctuations are expected to evolve as:
\begin{equation}
\label{nD_example_moment_eqn_mean}
\frac{d}{dt}\mathbb{E}[y_i] = w_i(\mathbf{a})\mathbb{E}[y_i] + a_i\sum\limits_{k=1}^{m}M_{ik}\mathbb{E}[y_k]
\end{equation}
or, in matrix form:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$\displaystyle
	\frac{d}{dt}\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	=
	\begin{bmatrix}
	(r + \sum\limits_{j=1}^{m}M_{1j}a_j + a_1M_{11}) & a_1M_{12} & a_1M_{13} & \dots & \dots & \dots & a_1M_{1m}\\
	a_2M_{21} & (r + \sum\limits_{j=1}^{m}M_{2j}a_j + a_2M_{22}) & a_2M_{23} & \dots & \dots & \dots & a_2M_{2m}\\
	\vdots &  & \ddots & &  & & \vdots\\
	a_{i}M_{i1} & a_iM_{i2} & a_iM_{i3} & \dots & (r + \sum\limits_{j=1}^{m}M_{ij}a_j + a_iM_{ii}) & \dots & a_iM_{im}\\
	\vdots &  &  & & & \ddots & \vdots\\
	a_mM_{m1} & a_mM_{m2} & a_mM_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}M_{mj}a_j + a_mM_{mm})
	\end{bmatrix}
	\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	$}
\end{equation}
The eigenvalues of the first matrix on the RHS will tell us whether the fixed point $\mathbb{E}[y_i] = 0 \ \forall \ i$ (the only fixed point of this system) is stable, or whether fluctuations are expected to grow (up to the point where the fluctuations are so large that the WNA is no longer valid). In the $m=2$ case,~\cite{tao_stochastic_2007} have shown that $\mathbb{E}[y_i] = 0 \ \forall \ i$ is a stable fixed point for this system iff the point $\mathbf{y}$ is an ESS (in the usual game-theoretic sense) for the matrix game defined by the payoff matrix $\mathbf{M}$.

@@

In Chapter \ref{chap_infD_processes}, we used various approximations to arrive at the linear functional Fokker-Planck equation
\begin{equation}
\label{functional_WNE_chap_4}
    \frac{\partial P}{\partial t}(\zeta,t) = \int\limits_{\mathcal{T}}\left(-\frac{\delta}{\delta \zeta(x)}\left\{\mathcal{D}_{\zeta}[\mathcal{A}^{-}](x)P(\zeta,t)\right\}+\frac{1}{2}\mathcal{A}^{+}(x|\psi)\frac{\delta^2}{\delta\zeta(x)^2}\{P(\zeta,t)\}\right)dx
\end{equation}
for describing stochastic fluctuations $\zeta$ from the deterministic solution obtained by solving \eqref{deterministic_traj}. Our goal is now to find a method to effectively detect and describe evolutionary branches (modes in trait space, corresponding to individual morphs) for this process. Following the methods used by Tim Rogers and colleagues for various special cases~\citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015}, we will do this in a general manner by measuring the autocorrelation of the distribution of the population over trait space, a task made easier by moving to Fourier space.
\myfig{0.8}{Media/4.3_fourier_fig.png}{\textbf{Schematic description of Fourier analysis}. A function $\phi(x)$ (shown in red) over the trait space can be decomposed as the sum of infinitely many Fourier modes (shown in blue) $\phi_k$. In the Fourier dual space, we can look at the peaks of each of these Fourier modes: The magnitude of $\phi_k$ tells us how much it contributes to the actual function of interest $\phi$.}{fig_Fourier}
Specifically, a convenient theorem due to Weiner and Khinchin relates the autocorrelation of a probability distribution to its power spectral density via Fourier transformation. This has been extensively used in spatial ecology, and we too will make use of it here by carrying out a basis expansion of our functions in the Fourier basis $\{e^{ikx}\}_{k\in\mathbb{Z}}$. If $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ is a translation-invariant\footnote{This is horrible nomenclature by the mathematicians. Though `invariant' is the conventional name for this concept, the intended meaning is not really invariant but `equivariant'. Formally, let $\mathcal{F}$ be a suitable function space of real valued functions. For any $c \in \mathbb{R}$, let $T_c: \mathcal{F} \to \mathcal{F}$ be the translation operator on this space, defined by $T_c[f(x)] = f(x+c)$. An operator $L: \mathcal{F} \to \mathcal{F}$ is said to be translation-invariant if it commutes with $T_c$ for every $c \in \mathbb{R}$, \emph{i.e.} $T_c[L[f]] = L[T_c[f]] \ \forall \ f \in \mathcal{F} \ \forall \ c \in \mathbb{R}$. This is not as restrictive as it initially sounds. For example, both the Laplacian operator and the convolution operator are linear and translation invariant} linear operator, then $\exp(ikx)$ acts as an eigenfunction, significantly simplifying the calculations. We therefore assume that $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ takes the form:
\begin{equation*}
 \mathcal{D}_{\zeta}[\mathcal{A}^-](x,t) = L[\zeta(x,t)]   
\end{equation*}
for a translation-invariant linear operator $L$ that only depends on $x$ and $t$. The presence of phenotypic clustering and polymorphisms can be analyzed by examining the power spectrum of $\Tilde{P}_{0}(\zeta,s)$ over the trait space.
